---
title: "Hierarchical modelling tutorial"
output: html_notebook
---

This notebook is a tutorial on Hierarchical modelling using Stan that accompanies the youtube video https://youtu.be/dNZQrcAjgXQ

```{r}
require(rstan)
set.seed(1) #set seed 
```

First we read in the data
```{r}
data = read.csv('/Users/maggielieu/Desktop/STAN/GITHUB/STAN_tutorials/Hierarchical/premiereleague.csv',col.names = c('Home','score1', 'score2', 'Away'), stringsAsFactors = FALSE)
```

```{r}
data
```

```{r}
ng = nrow(data)
cat('we have G =', ng, 'games \n')
```

```{r}
nt = length(unique(data$Home))
cat('we have T = ', nt, 'teams \n')
```

# Traditional method 

We will assume that the goals scored come from a poisson distribution
s1 | theta_g1 ~ Poisson(theta_g1)  #game g score by home team
s2 | theta_g2 ~ Poisson(theta_g2)  #game g score by away team

Assuming a log-linear random effect model
log(theta_g1) = home + att_ht + def_at
log(theta_g2) = att_at + def_ht

where home is a constant for the advantage for the team hosting the game
att and def are the attack and defence abilities of the teams where the indices at,ht correspond to the t=1-20 teams. 

priors we willl use for the attack and defence abilities are very wide, essentially the teams' performances are independent of each other
home ~ normal(0,0.0001)
att[t] ~ normal(0, 2)
def[t] ~ normal(0, 2)

Now convert team names for each match into numbers
```{r}
teams = unique(data$Home)
ht = unlist(sapply(1:ng, function(g) which(teams == data$Home[g])))
at = unlist(sapply(1:ng, function(g) which(teams == data$Away[g])))
```

```{r}
# we will save the last 5 games to predict
np=5
ngob = ng-np #number of games to fit
my_data = list(
  nt = nt, 
  ng = ngob,
  ht = ht[1:ngob], 
  at = at[1:ngob], 
  s1 = data$score1[1:ngob],
  s2 = data$score2[1:ngob],
  np = np,
  htnew = ht[(ngob+1):ng],
  atnew = at[(ngob+1):ng]
)
```

```{r}
nhfit = stan(file = 'non_hier_model.stan', data = my_data)
```
```{r}
print(nhfit)
```
`Plot the predicted scores of the last 5 matches
```{r, fig.height=5, fig.width=5}
nhparams = extract(nhfit)
pred_scores = c(colMeans(nhparams$s1new),colMeans(nhparams$s2new))
true_scores = c(data$score1[(ngob+1):ng],data$score2[(ngob+1):ng] )
plot(true_scores, pred_scores, xlim=c(0,5), ylim=c(0,5), pch=20, ylab='predicted scores', xlab='true scores')
abline(a=0,  b=1, lty='dashed')

pred_errors = c(sapply(1:np, function(x) sd(nhparams$s1new[,x])),sapply(1:np, function(x) sd(nhparams$s1new[,x])))
arrows(true_scores, pred_scores+pred_errors, true_scores, pred_scores-pred_errors, length = 0.05, angle = 90, code = 3, col=rgb(0,0,0,0.3))
```

We can also look at the attack/defense of the teams:
```{r}
attack = colMeans(nhparams$att)
defense = colMeans(nhparams$def)

plot(attack,defense,xlim=c(-0.4,1.1))
abline(h=0)
abline(v=0)
text(attack,defense, labels=teams, cex=0.7, pos=4)
```



Or we can assume all of the abilities are the same, or at least very similar by setting a very tight prior on the attack and defence abilities N(0,0.001), essentially restricting the teams to have the same performance 

```{r}
nhpoolfit = stan(file = 'non_hier_model_pool.stan', data = my_data)
```

Again we can plot the predicted scores of the last 5 matches, note now how the predicted scores are all the same.
```{r, fig.width=5, fig.height=5}
nhpoolparams = extract(nhpoolfit)
pred_scores = c(colMeans(nhpoolparams$s1new),colMeans(nhpoolparams$s2new))
pred_errors = c(sapply(1:np, function(x) sd(nhpoolparams$s1new[,x])),sapply(1:np, function(x) sd(nhpoolparams$s1new[,x])))
true_scores = c(data$score1[(ngob+1):ng],data$score2[(ngob+1):ng] )
plot(true_scores, pred_scores, xlim=c(0,4), ylim=c(0,4), pch=20, ylab='predicted scores', xlab='true scores')
abline(a=0,  b=1, lty='dashed')
arrows(true_scores, pred_scores+pred_errors, true_scores, pred_scores-pred_errors, length = 0.05, angle = 90, code = 3)
```

and similarly the teams' attack and defense abilities are all close to 0.
```{r}
attack = colMeans(nhpoolparams$att)
defense = colMeans(nhpoolparams$def)

plot(attack,defense,xlim=c(-0.4,1.1))
abline(h=0)
abline(v=0)
text(attack,defense, labels=teams, cex=0.7, pos=4)
```

# Hierarchical model
In a hierarchical model, the parameters of interest, in our case the attack and defense ability are drawn from the population distribution. 
att[t] ~ normal(mu_att, tau_att)
def[t] ~ normal(mu_def, tau_def)

Instead we define priors on the population, known as the hyperpriors.
mu_att = normal(0,0.0001)
tau_att = gamma(0.1,0.1)
mu_def = normal(0, 0.0001)
tau_def = gamma(0.1,0.1)

```{r}
hfit = stan(file = 'hier_model.stan', data = my_data)
```

```{r}
print(hfit)
```

```{r}
pairs(hfit, pars=c('mu_att', 'tau_att', 'mu_def', 'tau_def'))
```

```{r, fig.height=5, fig.width=5}
hparams = extract(hfit)
pred_scores = c(colMeans(hparams$s1new),colMeans(hparams$s2new))
pred_errors = c(sapply(1:np, function(x) sd(hparams$s1new[,x])),sapply(1:np, function(x) sd(hparams$s1new[,x])))
true_scores = c(data$score1[(ngob+1):ng],data$score2[(ngob+1):ng] )
plot(true_scores, pred_scores, xlim=c(0,5), ylim=c(0,5), pch=20, ylab='predicted scores', xlab='true scores')
abline(a=0,  b=1, lty='dashed')
arrows(true_scores, pred_scores+pred_errors, true_scores, pred_scores-pred_errors, length = 0.05, angle = 90, code = 3, rgb(0,0,0,0.3))
```


```{r}
attack = colMeans(hparams$att)
attacksd = sapply(1:nt, function(x) sd(hparams$att[,x]))
defense = colMeans(hparams$def)
defensesd = sapply(1:nt, function(x) sd(hparams$def[,x]))

plot(attack,defense, xlim=c(-0.4,1), ylim=c(-0.45,0.3), pch=20)
arrows(attack-attacksd, defense, attack+attacksd, defense, code=3, angle = 90, length = 0.04, col=rgb(0,0,0,0.2))
arrows(attack, defense-defensesd, attack, defense+defensesd, code=3, angle = 90, length = 0.04,col=rgb(0,0,0,0.2))
#abline(h=0)
#abline(v=0)
text(attack,defense, labels=teams, cex=0.7, adj=c(-0.05,-0.8) )
```













