---
title: "Linear_regression_I"
output: html_notebook
---

This is a tutorial on simple linear regression with Stan using the iris dataset. 

### First get the data
```{r}
data(iris) 
head(iris)
```


We only use the versicolor flowers, and define the covariates as the sepal length and the variates as the petal length. 
```{r}
versicolor = which(iris$Species == 'versicolor')
x = iris$Sepal.Length[versicolor]
y = iris$Petal.Length[versicolor] 
```

We prepare the data for Stan in a list format
```{r}
data = list(
	N = length(x),
	x = x,
	y = y
)
```

### Fit the data
```{r}
require(rstan)
fit = stan(file = 'model.stan',data=data)
```

```{r}
print(fit)
```

Extract the best fit parameters and visualise on the data
```{r}
plot(x,y)
params = extract(fit)
alpha = mean(params$alpha)
beta = mean(params$beta)
abline(a=alpha, b=beta)
```

### posterior intervals
The posterior draws can be used to obtain posterior intervals on the fit
```{r}
plot(x,y)
params = extract(fit)
alpha = mean(params$alpha)
beta = mean(params$beta)
abline(a=alpha, b=beta)
xr=seq(4,7.5,0.1)
yCI = sapply(xr, function(x) quantile(params$beta*x + params$alpha, probs=c(0.05,0.95) )) #95% quantiles 
lines(xr, yCI[1,], col='red')
lines(xr, yCI[2,], col='red')
```

# Simulated data
Simulated data from the posterior can be used as a sanity check
```{r}
plot(density(y), xlim=c(2,6.5), ylim=c(0,1.4))
for(i in 1:10){lines(density(params$y_sim[i,]), col='red')}
```

Using the simulated data, we should recover similar parameters to the real data. To check this we can re-run the model with a simulated dataset.
```{r}
y_new = params$y_sim[20, ] #20th simulated dataset
data_new = list(
	N = length(x),
	x = x,
	y = y_new
)
fit_new = stan(file ='model.stan', data=data_new)
```



Plot the posteriors on the parameters of the simulated and real data
```{r}
par(mfrow=c(1,2))
params_new=extract(fit_new)
plot(density(params$alpha))
lines(density(params_new$alpha),col='red')
plot(density(params$beta))
lines(density(params_new$beta),col='red')
```























